Fast and Flexible Binary Instrumentation for the Kernel

### 2013-12-07 VEE-14/paper.tex
Not much progress today. Decided to either start shaving off the wrapper stuff, or maybe I will talk about wrappers and probes uniformly (given their similarily). Also brought in Guardrail as a motivating example for adjusting the granularity of instrumentation based on an analysis tool's scope.

@@

\begin{table*}[ht!]
\scriptsize
\begin{tabularx}{\linewidth}{| c | >{\raggedright}m{0.3\linewidth} | X |}
\hline
\bf Scope & \bf Description & \bf Example Tool \\
\hline
Kernel & All kernel code is instrumented. & Granary's \toolname{cfg} tool constructs inter- and intra-procedural control flow graphs and records per-basic block and per-edge performance counters. This data is useful for debugging an performance analysis. Applying the \toolname{cfg} tool to the whole kernel allows us to perform profile-guided optimization of kernel code, including a form of de-virtualization. That is, most indirect control flow instructions (\texttt{call}, \texttt{jmp}) within the kernel target only one destination. Fast checks can be inlined into kernel code that use direct control transfer instructions when possible, and fall-back on indirect control transfer instructions when necessary.\\
\hline
Module & All code from a specific kernel module or driver is instrumented. Granary detaches when module code invokes kernel code, and re-attaches when kernel code invokes module code. All non-module kernel code executes with zero overhead. & TODO \\
\hline
Function & All code within a particular function is instrumented. Granary attaches when the function is invoked, and detaches when the function returns or invokes another function. & TODO \\
\hline
Basic Block & Individual basic blocks are instrumented on demand. & Granary's address watchpoints framework can be tuned to instrument only code accessing specific data. A general protection fault occurs when native code dereferences a  ``watched address". Granary can transparently handle this fault by returning execution to an instrumented version of the faulting code (where the instrumentation prevents such faults). Native execution can resume at the end of the basic block. \\
\hline
\end{tabularx}
\caption{\label{tab:what_kinds}Different kinds to instrumentation strategies supported by Granary.}
\end{table*}

\begin{figure}[t]
\captionsetup[subfloat]{width=0.45\columnwidth}
\begin{tabularx}{\columnwidth}{XX}%
\subfloat[Full-kernel instrumentation.]{\hspace{2.5em}\epsfig{file=diagrams/kernel.eps,width=0.125\textwidth}} & %
\subfloat[Module-only instrumentation]{\hspace{2.5em}\epsfig{file=diagrams/module.eps,width=0.125\textwidth}} \\
\subfloat[Module-only instrumentation, with a function specialized to run natively by using a module function wrapper.]{\hspace{2.5em}\epsfig{file=diagrams/module_wrapper.eps,width=0.125\textwidth}} & %
\subfloat[\label{fig:module_probe}Module-only instrumentation, with a kernel function specialized to also be instrumented by using a probe-based wrapper.]{\hspace{2.5em}\epsfig{file=diagrams/module_probe.eps,width=0.125\textwidth}}
\end{tabularx}
\caption{Example of using different granularities of instrumentation to selectively instrument or not instrument specific code.}
\end{figure}

@@

Granary allows tool developers to precisely specify the scope or granularity of instrumentation. For example, Granary can instrument all kernel code, specific module code, or only specific function code.

%what code should be instrumented. Example granularities of instrumentation include the whole kernel, only specific modules, only specific functions, or on-demand instrumentation based on events (e.g. interrupts or exceptions).

\subsection{Kernel}
Like DRK and btkernel, Granary can instrument the whole kernel. Instrumenting the whole kernel requires interposing on the system call entrypoint (x86-64 \texttt{MSR\_LSTAR} model-specific register) and on the interrupt vector entrypoints (stored in the interrupt descriptor table). For whole-kernel instrumentation, Granary boostraps by duplicating the the early entrypoint instructions that are responsible for switching execution onto a kernel stack, and then injects a jump to translate/instrumented kernel code after the stack switch. The translation process continues when a system call or interrupt occurs and follows the jump from duplicated into translated code. Granary only translate code that is known to operate on a kernel stack\footnote{Not all kernel code executes on a kernel stack. For example, when a user space application invokes the \texttt{syscall} instruction, control is transferred to the kernel, but no stack switch occurs. The kernel's early system call entrypoint code is responsible for switching execution onto a kernel stack.} so that instrumentation tools can safely save transient data in a reentrant way by spilling that data to the stack.

\subsection{Modules}\label{sec:module}

@@

\subsection{Functions}\label{sec:function_wrapper}

Granary can instrument or not instrument functions in several different ways. For example, \emph{probe-based} instrumentation allows Granary to selectively instrument only a particular function, while \emph{wrapper-based} instrumentation allows Granary to isolate a particular function that is invoked by instrumented code, and choose to instrument that function or run it natively (all while having the flexibility to arbitrarily inspect that function's arguments and return values in a type-safe way).

\subsubsection{Probe-based instrumentation}

@@

\subsubsection{Wrapper-based instrumentation}

Wrapper-based instrumentation makes use of the same wrapper capabilities used by module-only instrumentation (\Cref{sec:module}), but is available regardless of the instrumentation granularity. Wrapper-based instrumentation is similar to probe-based instrumentation, but permits more selectivity in terms of the context in which the function will be wrapped. That is, a function can be wrapped in one way when executed by instrumented kernel code, and in another way when executed by instrumented module code. Finer-grained selectivity can be achieved by having individual instrumentation policies manually decide how to wrap a given function.

\begin{figure}[t!]
\lstset{language=C, tabsize=2, stepnumber=1}
\begin{lstlisting}[basicstyle=\scriptsize\ttfamily]
MODULE_FUNCTION_WRAPPER(__kmalloc, (size, gfp), {
  void *addr = __kmalloc(size, gfp);
  if(is_valid_address(addr)) {
    add_watchpoint(addr, addr, size);
  }
  return addr;
})
\end{lstlisting}
\caption{Example module function wrapper for the \texttt{\_\_kmalloc} function. This wrapper adds an ``address watchpoint" to the address returned by the kernel's \texttt{\_\_kmalloc} memory allocator. Granary's module memory leak detector uses address watchpoints to track the liveness of module-allocated memory. Every access of a ``watched address" marks the associated memory as live within the current epoch. Memory that has not been live for several consecutive epochs is treated as a potential leak, and a is checked for reachability using a heap memory scanner.}
\label{fig:malloc_wrapper}
\end{figure}

Granary's module memory leak detector is one example application that uses the selectivity of wrappers (with respect to their calling context) to distinguish between module-allocated and kernel-allocated memory. \Cref{fig:malloc_wrapper} shows a module function wrapper that adds ``address watchpoints"  \cite{BehaveOrBeWatched} to module-allocated memory.

@@

\subsection{On-demand}

>>>>

\begin{table*}[ht!]
\scriptsize
\begin{tabularx}{\linewidth}{| l | >{\raggedright}p{0.3\linewidth} | X |}
\hline
\bf Granularity & \bf Description & \bf Example Tool \\
\hline
Kernel & All kernel code is instrumented. & Granary's \toolname{cfg} tool constructs inter- and intra-procedural control flow graphs and records per-basic block and per-edge performance counters. This data is useful for debugging an performance analysis. Applying the \toolname{cfg} tool to the whole kernel allows us to perform profile-guided optimization of kernel code, including a form of de-virtualization. That is, most indirect control flow instructions (\texttt{call}, \texttt{jmp}) within the kernel target only one destination. Fast checks can be inlined into kernel code that use direct control transfer instructions when possible, and fall-back on indirect control transfer instructions when necessary.\\
\hline
Module & All code from a specific kernel module or driver is instrumented. Granary detaches when module code invokes kernel code, and re-attaches when kernel code invokes module code. All non-module kernel code executes with zero overhead. & Granary's Guardrail tool records traces all instructions and memory operations performed by device driver code \cite{Guardrail}. A sideline thread analyzes the recorded traces to detect data race bugs and memory access bugs. \\
\hline
Function & All code within a particular function is instrumented. Granary attaches when the function is invoked, and detaches when the function returns or invokes another function. & TODO \\
\hline
Basic Block & Individual basic blocks are instrumented on demand. & Granary's address watchpoints framework can be tuned to instrument only code accessing specific data. A general protection fault occurs when native code dereferences a  ``watched address". Granary can transparently handle this fault by returning execution to an instrumented version of the faulting code (where the instrumentation prevents such faults). Native execution can resume at the end of the basic block. \\
\hline
\end{tabularx}
\caption{\label{tab:granularities}Different granularities of instrumentation supported by Granary.}
\end{table*}

@@

Some instrumentation tools have a limited scope. That is, they are made to solve a specific problem within a specific domain. For example, Granary's Guardrail tool is concerned with finding bugs in Linux device drivers \cite{Guardrail}. Guardrail records and analyzes traces of device driver code to find bugs. However, Guardrail's analysis can be heavyweight, and so limiting the analysis to only device driver code results in better performance. Granary allows Guardrail to set the granularity of instrumentation to only instrument a specific module, while leaving the rest of the kernel to execute without overhead.

%The recording and analysis phases of Guardrail are heavyweight, therefore, limiting

% Instrumenting the whole kernel to observer driver (module) execution is unnecessary and potentially very inefficient (especially if the tool is meant for online analysis). That is,

Guardrail is an example of how tools can choose an instrumentation granularity that better matches their scope and needs. That is, a Granary tool can choose to instrument some code, while leaving other code to execute natively. This is useful when the instrumentation in question has a high overhead or when only a subset of the kernel's code is of interest. \Cref{tab:granularities} provides a high-level overview of the different granularities of instrumentation supported by Granary. Granary tools, however, are not restricted to only the Granularities listed in the table.

%instrument all kernel code, specific module code, or only specific function code.
%what code should be instrumented. Example granularities of instrumentation include the whole kernel, only specific modules, only specific functions, or on-demand instrumentation based on events (e.g. interrupts or exceptions).

\subsection{Instrumenting All Kernel Code}
Like DRK and btkernel, Granary can instrument the whole kernel. Instrumenting the whole kernel requires interposing on the system call entrypoint (x86-64 \texttt{MSR\_LSTAR} model-specific register) and on the interrupt vector entrypoints (stored in the interrupt descriptor table). For whole-kernel instrumentation, Granary boostraps by duplicating the the early entrypoint instructions that are responsible for switching execution onto a kernel stack, and then injects a jump to translate/instrumented kernel code after the stack switch. The translation process continues when a system call or interrupt occurs and follows the jump from duplicated into translated code. Granary only translate code that is known to operate on a kernel stack\footnote{Not all kernel code executes on a kernel stack. For example, when a user space application invokes the \texttt{syscall} instruction, control is transferred to the kernel, but no stack switch occurs. The kernel's early system call entrypoint code is responsible for switching execution onto a kernel stack.} so that instrumentation tools can safely save transient data in a reentrant way by spilling that data to the stack.

\begin{figure}[t]
\captionsetup[subfloat]{width=0.45\columnwidth}
\begin{tabularx}{\columnwidth}{XX}%
\subfloat[Full-kernel instrumentation.]{\hspace{2.5em}\epsfig{file=diagrams/kernel.eps,width=0.125\textwidth}} & %
\subfloat[Module-only instrumentation]{\hspace{2.5em}\epsfig{file=diagrams/module.eps,width=0.125\textwidth}} \\
\subfloat[Module-only instrumentation, with a function specialized to run natively by using a module function wrapper.]{\hspace{2.5em}\epsfig{file=diagrams/module_wrapper.eps,width=0.125\textwidth}} & %
\subfloat[\label{fig:module_probe}Module-only instrumentation, with a kernel function specialized to also be instrumented by using a probe-based wrapper.]{\hspace{2.5em}\epsfig{file=diagrams/module_probe.eps,width=0.125\textwidth}}
\end{tabularx}
\caption{Visual example of using different granularities of instrumentation to selectively instrument or not instrument specific code.}
\end{figure}

\subsection{Instrumenting Only Module Code}\label{sec:module}

@@

@@

\subsection{Instrumenting Specific Functions}\label{sec:function_wrapper}

%Granary can instrument or not instrument functions in several different ways. For example, \emph{probe-based} instrumentation allows Granary to selectively instrument only a particular function, while \emph{wrapper-based} instrumentation allows Granary to isolate a particular function that is invoked by instrumented code, and choose to instrument that function or run it natively (all while having the flexibility to arbitrarily inspect that function's arguments and return values in a type-safe way).

%\subsubsection{Probe-based instrumentation}

@@

%\subsubsection{Wrapper-based instrumentation}
%
%Wrapper-based instrumentation makes use of the same wrapper capabilities used by module-only instrumentation (\Cref{sec:module}), but is available regardless of the instrumentation granularity. Wrapper-based instrumentation is similar to probe-based instrumentation, but permits more selectivity in terms of the context in which the function will be wrapped. That is, a function can be wrapped in one way when executed by instrumented kernel code, and in another way when executed by instrumented module code. Finer-grained selectivity can be achieved by having individual instrumentation policies manually decide how to wrap a given function.
%
%\begin{figure}[t!]
%\lstset{language=C, tabsize=2, stepnumber=1}
%\begin{lstlisting}[basicstyle=\scriptsize\ttfamily]
%MODULE_FUNCTION_WRAPPER(__kmalloc, (size, gfp), {
%  void *addr = __kmalloc(size, gfp);
%  if(is_valid_address(addr)) {
%    add_watchpoint(addr, addr, size);
%  }
%  return addr;
%})
%\end{lstlisting}
%\caption{Example module function wrapper for the \texttt{\_\_kmalloc} function. This wrapper adds an ``address watchpoint" to the address returned by the kernel's \texttt{\_\_kmalloc} memory allocator. Granary's module memory leak detector uses address watchpoints to track the liveness of module-allocated memory. Every access of a ``watched address" marks the associated memory as live within the current epoch. Memory that has not been live for several consecutive epochs is treated as a potential leak, and a is checked for reachability using a heap memory scanner.}
%\label{fig:malloc_wrapper}
%\end{figure}
%
%Granary's module memory leak detector is one example application that uses the selectivity of wrappers (with respect to their calling context) to distinguish between module-allocated and kernel-allocated memory. \Cref{fig:malloc_wrapper} shows a module function wrapper that adds ``address watchpoints"  \cite{BehaveOrBeWatched} to module-allocated memory.

@@

\subsection{On-Demand Instrumentation}
